---
title: "Blocking in Experiments"
subtitle: |
  | Additional Topic 
  | Math 215
author: "Yurk"
format: 
  revealjs:
    theme: beige
editor: source
---

```{r}
#| include: false
#| echo: false

library(tidyverse)
library(infer)
library(openintro)
library(broom)
library(HH)
```

## Experiments with two factors

- A **factor** is a categorical variable that we think may explain some of the variability in the response variable
- A **treatment factor** is a factor that we want to evaluate to see if there is an effect
- A **blocking factor** is a factor that we believe will have an effect
- The treatment factor is the factor of interest
- The blocking factor is used to reduce the unexplained variability in the response

---

- An experiment with two factors may have two treatment factors (more on this later)
- Or it may have a treatment factor and a blocking factor

## Experimental Designs

**Randomize complete block design (RCBD)**

- 1 treatment factor with $t$ levels, 1 blocking factor with $b$ levels
- Each block has exactly $t$ experimental units (one for each treatment level)
- Experimental units in same block expected to respond similarly if treated similarly
- **Matched pair design**: $t=2$
- **Repeated measures design**: Each block is a single individual, subject to all $t$ treatments
  
---
  
**Factorial design** (more on this later)

- More general than RCBD
- Two or more factors
- One or more observations per cell (replication)

## Strawberries

```{r}
#| include: false
#| echo: false

strawberries <- read_tsv("data/StrawberryStorageRCBD.txt")
```

-   Does changing the air in which strawberries are stored affect the firmness of the berries?
- Data from Smith and Skog (1992) [^1]
- Inspired by an example from [Tintle et al. (2020)](http://www.isi-stats.com/isi2/)

[^1]: Smith, R.B., Skog, L.J. 1992. Postharvest carbon dioxide treatment enhances firmness of several cultivars of strawberry, Horticulture Science 27, Pp. 420-421

---

- Response: `Firmness` (force in N to pierce berry)
- Treatment factor: `Storage` with 3 levels ($t=3$):

  - "Control" (not stored)
  - "Air" (21% O<sub>2</sub>, 0.04% CO<sub>2</sub>)
  - "ModifiedAir" (15% CO<sub>2</sub>, 18% O<sub>2</sub>)
  
- Blocking factor: `Variety` with 5 levels ($b=5$): "Allstar", "Bounty", "Kent", "Selva", "Vesper"
- Stored berries were stored at $0.5^{\circ}$C for two days
- Three clamshells of each variety randomly assigned to treatments

## Statistical Model

$$y_{ij}=\mu+\tau_i+\rho_j+\varepsilon_{ij}$$

- $\mu$ is the overall mean
- $\tau_i$ is the differential effect of treatment $i$
- $\rho_j$ is the differential effect of block $j$
- $\varepsilon_{ij}\sim N(0,\sigma^2)$ represents the error
- This is an **additive model**: the treatment is assumed to have the same effect in each block

## Hypothesis Test

- We expect firmness to vary according to variety
- However, we are interested in the effect of the treatment (air type)
- We conduct a hypothesis test for the treatment variable:

  - $H_0: \tau_1 = \tau_2 = \tau_3 = 0$
  - $H_A:$ At least one $\tau_i$ is different

## EDA

::: panel-tabset
### Storage

```{r}
#| include: TRUE
#| echo: FALSE
#| fig-cap: "Strawberry firmness for different storage conditions."

strawberries |>
  ggplot(aes(Storage, Firmness)) +
  geom_boxplot() +
  theme_minimal()
```

### Variety

```{r}
#| include: TRUE
#| echo: FALSE
#| fig-cap: "Firmness for different strawberry varieities."

strawberries |>
  ggplot(aes(Variety, Firmness)) +
  geom_boxplot() +
  theme_minimal()
```

### Both

```{r}
#| include: TRUE
#| echo: FALSE
#| fig-cap: "Firmness for different strawberry varieities with different storage conditions."

strawberries |>
  ggplot(aes(Variety, Firmness, color = Storage, shape = Storage)) +
  geom_point(size = 4) +
  theme_minimal()
```
:::

## ANOVA Table (Not acconting for Variety)

- First we consider ANOVA without accounting for `Variety`
- We are unable to detect a treatment effect
- Much of the variability in firmness is due to differences between varieties and is left unexplained

```{r}
#| include: TRUE
#| echo: TRUE

lm(Firmness ~ Storage, data = strawberries) |>
  anova() |>
  tidy()
```

## ANOVA Table (Accounting for variety)

- This time we account for `Variety`
- Now the treatment effect is apparent
- We reject $H_0$ and conclude that there is convincing evidence of a treatment effect.

```{r}
#| include: TRUE
#| echo: TRUE

lm(Firmness ~ Variety + Storage, data = strawberries) |>
  anova() |>
  tidy()
```

## Pairwise Comparisons

- We can follow up with pairwise comparisons between different treatment levels
- Adjust for multiple comparisons

```{r}
#| include: TRUE
#| echo: TRUE

aov(Firmness ~ Variety + Storage, data = strawberries) |>
  TukeyHSD("Storage")
```

---

There is a significant difference between the mean firmness for "Modified Air" and "Air" and between "Modified Air" and "Control"

## Scope of Inference

- Because this was an experiment, we can conclude that the difference in storage method caused the difference in firmness
- Storing strawberries in air enriched in CO<sub>2</sub> increases firmness compared to not storing the berries at all or storing them in normal air

---

- Choosing varieties with large variability in firmness (large variation between blocks) broadens the scope of inference
- For example, we could have controlled unexplained variability by focusing on a single variety
- Then our conclusions would be limited to that single variety