---
title: "Hypothesis Testing with Randomization"
subtitle: |
  | IMS1 Ch. 11 
  | Math 215
author: "Yurk"
format: 
  revealjs:
    theme: beige
    df-print: paged
editor: visual
---

## Flavor Preferences

<!-- Useful: https://quarto.org/docs/presentations/revealjs/ -->

<!-- this, too: https://quarto.org/docs/authoring/markdown-basics.html -->

```{r}
#| include: false
#| echo: false

library(tidyverse)
library(infer)
library(openintro)
```

```{r}
#| include: false
#| echo: false

soda <- data.frame(
  location = c(
    rep("East", 28), rep("West", 19),
    rep("East", 6), rep("West", 7)
  ),
  drink = c(rep("Cola", 47), rep("Orange", 13))
) |>
  arrange(location)
```

-   **Research question:** Do people on the East Coast have a higher preference for cola than people on the West Coast?
-   `soda` dataset
-   2 variables
    -   *location*: East or West
    -   *drink* preference: Orange or Cola
-   60 individuals (43 from East, 26 from West)

## Results (EDA)

::: panel-tabset
### Counts

```{r}
#| include: false
#| echo: false

opportunity_cost |>
  count(group, decision) |>
  pivot_wider(names_from = decision, values_from = n) |>
  mutate(total = `buy video` + `not buy video`)
```

| location | Cola | Orange | total |
|----------|:----:|:------:|------:|
| East     |  28  |   6    |    34 |
| West     |  19  |   7    |    26 |
| total    |  47  |   13   |    60 |

### Bar Plot

```{r}
#| include: true
#| echo: false
#| fig-cap: "Standardized barplot showing proportions of drink preferences"

soda |>
  ggplot(aes(x = location, fill = drink)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("#542a18", "#ffa500")) +
  theme_minimal()
```
:::

## Difference in proportions

```{r}
#| include: false
#| echo: false

obs_dif_prop <- soda |>
  group_by(location) |>
  summarize(prop = mean(drink == "Cola"))  |>
  summarize(dif = -diff(prop)) |> #East - West
  pull()
```

-   Success: drink = Cola
-   Statistic of interest: difference in proportions $$\hat{p}_E-\hat{p}_W$$
-   Observed difference: $$\frac{28}{34}-\frac{19}{26}=0.09276$$

## Hypothesis Test

-   From the sample it appears that there is a stronger preference for cola on the East Coast
-   It may be that there is no real difference in preference in the population, and the observed difference is not surprising when selecting a sample of this size from the population
-   A **hypothesis test** states these two possibilities formally as hypotheses then weighs them against each other using the results from the sample as evidence

## Hypotheses

-   The **null hypothesis**, denoted $H_0$, represents a skeptical perspective or a claim of no difference
-   The **alternative hypothesis**, denote $H_A$, represents an alternative claim of difference.
-   As statisticians, we usually establish hypotheses *before* viewing the data in order to avoid bias

------------------------------------------------------------------------

::: columns
::: {.column width="50%"}
In words:

|     $H_0:$ Location has no
|     effect on preference for
|     cola over orange soda.

|     $H_A:$ There is a higher
|     preference for cola
|     over orange soda on the
|     East Coast than on the
|     West Coast.
:::

::: {.column width="50%"}
In symbols:

|     $H_0: p_E - p_W = 0$
|     $H_A: p_E - p_W > 0$
:::
:::

## Null Distribution

-   We **test the null hypothesis** by comparing the observed value of the statistic to a **null distribution**
-   If the null hypothesis is *true* and we select different samples of the same size from the population, we would expect the value of the statistic to vary between samples
-   The null distribution is the distribution that describes those values
-   It is an example of a **sampling distribution** (distribution of a statistic)

## Null Distribution Using Random Permutation

-   Suppose that I suspect Hope students that sit in the front of class had a higher high school GPA than students that sit in the back
-   I ask each of you to write your high school GPA on a sheet of paper and I calculate the difference in mean GPA for students in the front and in the back
-   I want to know how that difference compares to differences I would measure if there is no difference

------------------------------------------------------------------------

-   The GPAs I collected is my best picture of what the distribution of GPAs is like at Hope
-   To simulate the null hypothesis being true (no difference between front and back), I could mix up your GPAs and hand them back to you
-   Then I could collect them again and measure the difference in means between front and back
-   If I do this many times it will give me a good idea of what the differences would look like if the null hypothesis is true (the null distribution)

------------------------------------------------------------------------

-   Mixing up the values of the response variable as in the GPA example is called **random permutation**
-   I can use random permutation to create a null distribution
-   Usually we will do this with a computer, because we want to calculate the statistic for 1,000 or 10,000 random permutations

------------------------------------------------------------------------

Here is the original soda data with 5 random permutations.

```{r}
#| include: true
#| echo: false
set.seed(8675309)

library(infer) 
soda_perm5 <- soda |> 
  specify(drink ~ location, success = "Cola") |>
  hypothesize("independence") |>
  generate(reps = 5, type = "permute") |>
  ungroup() |>
  mutate(id = rep(1:60, 5)) |>
  pivot_wider(id_cols = id, names_from = replicate, values_from = drink, names_prefix = "randPerm")

soda_perm5 <- soda |>
  mutate(id = 1:60) |>
  full_join(soda_perm5) |>
  relocate(id)

soda_perm5
```

------------------------------------------------------------------------

-   Now let's simulate 100 samples assuming true null hypothesis
-   We'll calculate a difference in proportions for each permutation
-   Use `infer` package

```{r}
#| include: true
#| echo: true
set.seed(8675309)

library(infer)
soda_perm <- soda |> 
  specify(drink ~ location, success = "Cola") |>
  hypothesize("independence") |>
  generate(reps = 100, type = "permute") |>
  calculate(stat = "diff in props", order = c("East", "West"))
```

------------------------------------------------------------------------

```{r}
#| include: true
#| echo: false
#| fig-cap: "Dot plot of 100 differences in randomized proportions (null distribution), showing observed difference as dashed vertical line."

soda_perm |>
  ggplot(aes(x = stat)) +
  geom_dotplot(dotsize = 0.3, stackratio = 1.5) +
  geom_vline(xintercept = obs_dif_prop, color = "red", linetype = "dashed") +
  labs(title = "100 differences in randomized proportions",
       x = "difference in randomized proportions of people who prefer cola (East - West)") +
  scale_y_continuous(NULL, breaks = NULL) +
  theme_minimal()
```

## p-Value

-   To test the null hypothesis ($p_E-p_W = 0$) we consider how probable it would be to get a difference in proportions that is at least as large as the observed difference if $H_0$ is true
-   This probability is called a **p-value**
-   We use the null distribution to calculate the p-value

------------------------------------------------------------------------

```{r}
#| include: false
#| echo: false

soda_perm |>
  summarize(n_extreme = sum(stat >= obs_dif_prop),
          pval = mean(stat >= obs_dif_prop))
```

There are 28 differences in randomization proportions that are greater than or equal to the observed value (0.09276). So we estimate the p-value to be 28/100 = 0.28.

```{r}
#| include: true
#| echo: false
#| fig-cap: "Dot plot of 100 differences in randomized proportions (null distribution), showing observed difference as dashed vertical line."

soda_perm |>
  ggplot(aes(x = stat)) +
  geom_dotplot(dotsize = 0.3, stackratio = 1.5) +
  geom_vline(xintercept = obs_dif_prop, color = "red", linetype = "dashed") +
  labs(title = "100 differences in randomized proportions",
       x = "difference in randomized proportions of people who prefer cola (East - West)") +
  scale_y_continuous(NULL, breaks = NULL) +
  theme_minimal()
```

## Significance Level

-   *Before* we conduct a study, we define a **significance level**, denoted $\alpha$
-   We decide that in order to reject the null hypothesis as false, the p-value must be less than $\alpha$
-   The significance level is the *standard of evidence* we will use to judge the null hypothesis
-   We presume the null hypothesis is true, but we are willing to reject it if the evidence against it is strong enough (the p-value is less than $\alpha$)

------------------------------------------------------------------------

-   Typical values for $\alpha$ are 0.05 and 0.01
-   Sometimes other values are used
-   Unless otherwise noted, we will always use $\alpha = 0.05$

## Conclusion

-   In the soda example, the observed difference in proportions ($\hat{p}_E-\hat{p}_W = 0.09276$) does not allow us to reject the null hypothesis (p = 0.28) at the $\alpha = .05$ significance level.
-   The difference in the proportions is not **statistically significant**
-   This means that it is **plausible** that there is no difference in the proportions of people who prefer cola to orange soda between the East and West Coast.
