---
title: "Decision Errors"
subtitle: |
  | IMS1 Ch. 14 
  | Math 215
author: "Yurk"
format: 
  revealjs:
    theme: beige
editor: source
---

## CPR Study

<!-- Useful: https://quarto.org/docs/presentations/revealjs/ -->

<!-- this, too: https://quarto.org/docs/authoring/markdown-basics.html -->

```{r}
#| include: false
#| echo: false

library(tidyverse)
library(infer)
library(openintro)
```

-   **Research question:** Do blood thinners affect survival rate in heart attack patients that have received CPR?
-   `cpr` [^1] dataset
-   2 variables
    -   *group*: treatment (received blood thinner) or control (did not)
    -   *outcome*: died or survived (for at least 24 hours)
-   90 patients (40 treatment, 50 control, randomly assigned)

[^1]: `cpr` is from the *openintro* package

## Hypotheses

-   Blood thinners can be administered to treat a clot that is causing a heart attack
-   CPR can cause internal injuries
-   Blood thinners can make it more difficult for these injuries to heal
-   Do blood thinners affect survival in a positive or negative way?
-   Alternative hypothesis reflects the fact that we don't have expectations about the direction of the relationship

------------------------------------------------------------------------

Two-sided hypothesis test

-   $H_0$: Blood thinners do not affect survival rate. $p_T-p_C = 0$

-   $H_A$: Blood thinners affect survival rate. $p_T-p_C \neq 0$

Example of one-sided hypothesis test

-   $H_0$: Blood thinners do not affect survival rate. $p_T-p_C = 0$

-   $H_A$: Blood thinners increase survival rate. $p_T-p_C > 0$

## Results (EDA)

::: panel-tabset
### Counts

```{r}
#| include: false
#| echo: false

cpr |>
  count(group, outcome) |>
  pivot_wider(names_from = outcome, values_from = n) |>
  mutate(total = died + survived)
```

| group     | died | survived | total |
|-----------|:----:|:--------:|------:|
| control   |  39  |    11    |    50 |
| treatment |  26  |    14    |    40 |
| total     |  65  |    25    |    90 |

### Bar Plot

```{r}
#| include: true
#| echo: false
#| fig-cap: "Standardized barplot showing proportions of students that buy or do not buy"

cpr |>
  ggplot(aes(x = group, fill = outcome)) +
  geom_bar(position = "fill") +
  labs(y = "proportion") +
  theme_minimal()
```
:::

## Difference in proportions

```{r}
#| include: false
#| echo: false

cpr |>
  group_by(group) |>
  summarize(prop = mean(outcome == "survived"))  |>
  summarize(dif = diff(prop)) #treatment - control
```

-   Success: outcome = "survived"
-   Statistic of interest: difference in proportions $$\hat{p}_T-\hat{p}_C$$
-   Observed difference: $$\frac{14}{40}-\frac{11}{50}=0.13$$

## Hypothesis test

```{r}
#| include: false
#| echo: false

cpr |>
  group_by(group) |>
  summarize(n = n(), prop = mean(outcome == "survived")) |>
  mutate(varpi = prop*(1-prop)/n) |>
  summarize(se = sqrt(sum(varpi)))
```

-   Technical conditions met for using a normal approximation
-   SE = 0.0955 (more on estimating this in Ch. 17)
-   Null distribution: $N(0, 0.0955)$
-   We will use significance level $\alpha=0.05$

------------------------------------------------------------------------

```{r}
#| include: true
#| echo: false
#| fig-cap: "Density function for N(0, 0.0955) with tails shaded beyond 0.13."

normTail(m = 0, s = 0.0995, L = -0.13, U = 0.13)
```

-   p-value is probability that statistic is at least as extreme as observed value (0.13)
-   2-sided test: at least as extreme means $\geq 0.13$ or $\leq -0.13$

------------------------------------------------------------------------

-   Since the null distribution is symmetric, the p-value will be twice as large for a 2-sided test

```{r}
#| include: true
#| echo: true

2 * pnorm(-0.13, mean = 0, sd = 0.0995)
```

-   Since the p-value is greater than 0.05, we do not reject the null hypothesis
-   It is plausible that there is no difference in survival rates between the two groups

## Decision Errors

-   There are two types of errors we can make in a hypothesis test
-   **Type 1 Error** occurs if we conclude that the null hypothesis is false when it is not (false alarm)
-   **Type 2 Error** occurs if we fail to reject the null hypothesis even though it is false (missed opportunity)

## Controlling Type 1 Errors

-   Type 1 Error is typically considered more severe
-   The probability of making a type 1 error (assuming the null is true) is equal to the significance level
-   Can reduce $\alpha$ to make type 1 errors less likely

## Controlling Type 2 Errors

-   There is a trade-off between the two types of errors
-   Decreasing probability of type 1, increases the probability of type 2
-   **Power** is the probability of rejecting the null hypothesis if the alternative is true
-   Higher power reduces the chance of making a type 2 error
-   Power is related to effect size (easier to detect larger effects), sample size (larger sample results in more power), among other things
