---
title: "Inference for a Single Proportion"
subtitle: |
  | IMS1 Ch. 16 
  | Math 219
author: "Yurk"
format: 
  revealjs:
    theme: beige
editor: source
---

## Medical Consultants

<!-- Useful: https://quarto.org/docs/presentations/revealjs/ -->

<!-- this, too: https://quarto.org/docs/authoring/markdown-basics.html -->

```{r}
#| include: false
#| echo: false

library(tidyverse)
library(infer)
library(openintro)
```

-   Some organ donors work with a medical consultant who helps them throughout the process
-   The average complication rate for liver donor surgeries in the United States is about 10%
-   One consultant claims she has low rate of complications compared to national average. Is her claim supported?
-   Let $p$ be the consultant's long-run complication rate

------------------------------------------------------------------------

```{r}
#| include: false
#| echo: false

consult <- tibble(complication = c(rep("yes", 3), rep("no", 59)))
```

Hypotheses:

-   $H_0: p = 0.1$
-   $H_A: p < 0.1$

Data:

-   `consult` dataset
-   She has served as a consultant for 62 liver donor surgeries
    -   3 (4.8%) resulted in complications
    -   $\hat{p} = 0.048$

## Payday Loan Regulations

-   Borrowers use payday loans to get a cash advance before their next payday
-   Borrower writes a check for loan amount + service fee
-   Lender holds check until borrower's payday
-   Very high APR equivalent (often over 300%)
-   Some borrowers take out second loan to pay of first, and so on

------------------------------------------------------------------------

-   Michigan already has a law that limits the number of payday loans a borrower can hold (2)
-   Do most payday borrowers support additional regulation that would require payday lenders to do a credit check?
-   Let $p$ be the proportion of payday borrowers in MI that support additional regulation.

------------------------------------------------------------------------

```{r}
#| include: false
#| echo: false

payday <- tibble(support = c(rep("yes", 424), rep("no", 402)))
```

Hypotheses:

-   $H_0: p = 0.5$
-   $H_A: p > 0.5$

Data:

-   `payday` dataset
-   Researchers selected a random sample of 826 payday borrowers
    -   424 (51.3%) said they would support a regulation
    -   $\hat{p}=0.513$

## Mathematical Model for a Proportion

-   We have learned that if certain conditions are met we can use a mathematical model to make inferences about a population
-   There is a version of the Central Limit Theorem for a single proportion

------------------------------------------------------------------------

::: callout-note
## Sampling distribution of $\hat{p}$

The sampling distribution of $\hat{p}$ based on a sample of size $n$ from a poplation with true proportion $p$ will be approximately normal with mean $p$ and standard error $$SE=\sqrt{\frac{p(1-p)}{n}}$$\
if the following technical conditions are met:

1.  independent observations (e.g., observations from SRS)
2.  (**success-failure condition**) at least 10 expected successes and at least 10 expected failures. (i.e., $np\geq 10$ and $n(1-p)\geq 10$)
:::

## Checking Technical Conditions

Consultant study

-   The success-failure condition is not met. Under the null hypothesis, we expect $62\times 0.1 = 6.2$ complications (less than 10)
-   Cannot model null distribution using a normal distribution
-   Use randomization instead (**parametric bootstrap simulation**)

------------------------------------------------------------------------

Payday study

-   The success-failure condition is met. Under the null hypothesis, we expect $0.5\times 826 = 413$ people to support the legislation and $(1-0.5)\times 826 = 413$ to not support the legislation.
-   It is appropriate to model the null distribution using a normal distribution

## Hypothesis Test Using Normal Approximation

-   Let $p_0$ be the proportion under the null hypothesis
-   We will use the **Z score** as the test statistic $$Z = \frac{\hat{p}-p_0}{SE(\hat{p})}=\frac{\hat{p}-p_0}{\sqrt{p_0(1-p_0)/n}}$$
-   Recall that if $\hat{p}$ is normally distributed, then $Z$ has a standard normal distribution, $N(0,1)$

------------------------------------------------------------------------

```{r}
#| include: false
#| echo: false

p0 <- 0.5
phat <- 424/826
n <- 826
Z <- (phat - p0)/sqrt(p0*(1-p0)/n)
```

-   For the payday study $p_0=0.5$, and $\hat{p}=0.513$, so $$Z = \frac{0.513 - 0.5}{\sqrt{0.5\cdot(1-0.5)/826}}=0.765$$
-   The p-value is the probability that we would obtain a $Z$ score at least as large as 0.765 if the null hypothesis is true
-   We compute the p-value by finding the area under the density curve for $N(0,1)$ that is beyond 0.765

```{r}
#| include: false
#| echo: false

p0 <- 0.5
phat <- 424/826
n <- 826
Z <- (phat - p0)/sqrt(p0*(1-p0)/n)
```

------------------------------------------------------------------------

```{r}
#| include: true
#| echo: false
#| fig-cap: "Normal model, $N(0,1)$. P-value is area of shaded region."

normTail(m = 0, s = 1, U = Z)
```

```{r}
#| include: true
#| echo: true

1 - pnorm(0.765, mean = 0, sd = 1)
```

------------------------------------------------------------------------

-   We are unable to reject the null hypothesis (p-value = 0.22)
-   Note that we cannot claim that 50% of payday borrowers support the new legislation (we cannot accept the null hypothesis)
-   However, 50% is a plausible value for the parameter

## Hypothesis Test Using Randomization

-   In the consultant study we cannot use a normal model for the null distribution
-   However, we can use **parametric bootstrap simulation** to approximate the null distribution
-   We simulate 1,000 random samples of 62 liver donors from a population in which the null hypothesis is true (10% complication rate)

------------------------------------------------------------------------

Parametric bootstrap simulation is equivalent to the following physical simulation:

-   For each donor simulate the outcome by spinning a spinner with 10% of the area representing "complication" and 90% representing "no complication"
-   For each sample, spin the spinner 62 times and record the proportion of complications in the sample
-   Repeat to obtain proportions for 1,000 simulated samples

------------------------------------------------------------------------

We can do the bootstrapping using the `infer` package

```{r}
#| include: true
#| echo: true

set.seed(8675309)

library(infer)
consult_boot <- consult |>
  specify(response = complication, success = "yes") |>
  hypothesize(null = "point", p = 0.1) |>
  generate(reps = 1000, type = "draw") |>
  calculate(stat = "prop")
```

------------------------------------------------------------------------

```{r}
#| include: true
#| echo: false
#| fig-cap: "Approximate null distribution with observed proportion of surgeries with complications (0.048) inticated by dashed vertical line."

consult_boot |>
  ggplot(aes(stat)) +
  geom_histogram(bins = 15) +
  geom_vline(xintercept = 3/62, 
             color = "red", linewidth = 2, linetype = "dashed") +
  labs(x = "bootstrapped proportion of surgeries with complications",
       title = "1,000 parametric bootstrapped proportions (p = 0.1)") +
  theme_minimal()
```

------------------------------------------------------------------------

-   The p-value is approximated by the proportion of bootstrapped proportions that are at least as extreme as the observed proportion ($\leq 0.048$)
-   With a p-value of 0.11 we are unable to reject the null hypothesis.
-   It is plausible that the consultant has the same complication rate as the national average of 10%

```{r}
#| include: true
#| echo: true

consult_boot |>
  summarize(n_extreme = sum(stat <= 3/62),
            p_val = mean(stat <= 3/62))
```

## Confidence Interval

-   We can also use a normal distribution to find a confidence interval if technical conditions are met
-   Earlier we used $p_0$ as the mean and in the computation of SE, because we were trying to approximate the null distribution
-   A confidence interval estimates the value of the parameter
-   The best point-estimate we have is the sample proportion $\hat{p}$, so we use that as the mean and in the computation of SE

## Checking Conditions for CI

-   The success-failure condition is easier to check in this situation.
-   $n\hat{p}$ is the number of observed success, and $n(1-\hat{p})$ is the number of observed failures.
-   We just need to check if there were 10 successes and 10 failures in the sample.

------------------------------------------------------------------------

Consultant study

-   The success-failure condition is not met. There were 3 successes and 59 failures
-   Cannot use normal approximation to find a CI
-   Use randomization instead (**bootstrap** as in Chapter 12)

------------------------------------------------------------------------

Payday study

-   The success-failure condition is met. There were 424 successes and 402 failures
-   It is appropriate to use a normal approximation to find a CI

## Confidence Interval Using a Normal Approximation

-   If a normal approximation is appropriate, a confidence interval for a proportion can be written as $$\hat{p}\pm z^{\ast}\times SE$$
-   SE is estimated using $$SE\approx\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$
-   $z^{\ast}$ is determined by the confidence level (e.g., 1.96 for 95%, 2.58 for 99%)

------------------------------------------------------------------------

```{r}
#| include: false
#| echo: false

n <- 826
phat <- 424/826
se <- sqrt(phat*(1-phat)/n)

cil <- phat - 1.96*se
ciu <- phat + 1.96*se
```

-   The standard error for the proportion of borrowers that support the new regulation is $$SE \approx \sqrt{\frac{0.513\cdot(1-0.513)}{826}}=0.0174$$
-   The 95% confidence interval is $$0.513\pm1.96\cdot0.0174 = 0.513\pm0.034$$
-   We are 95% confident that the proportion of payday borrowers that support the new regulation is between 0.479 and 0.547

## Confidence Interval Using a Bootstrapping

-   We use bootstrapping to find a 95% confidence interval for the complication rate for the medical consultant
-   This time we take repeated samples (with replacement) from our original sample

```{r}
#| include: true
#| echo: true

consult_boot_ci <- consult |>
  specify(response = complication, success = "yes") |>
  generate(reps = 1000, type = "bootstrap") |>
  calculate(stat = "prop")
```

------------------------------------------------------------------------

-   The 95 % bootstrap confidence interval is obtained by calculating the 2.5% and 97.5% percentiles for the bootstrapped statistics.
-   We are 95% confident that the consultant's long-run complication rate is between 0 and 0.113

```{r}
#| include: true
#| echo: true

consult_boot_ci |>
  summarize(ci_lo = quantile(stat, 0.025),
            ci_hi = quantile(stat, 0.975))
```
