---
title: "Comparing Two Proportions"
subtitle: |
  | IMS1 Ch. 17 
  | Math 219
author: "Yurk"
format: 
  revealjs:
    theme: beige
editor: source
---

## CPR Study

<!-- Useful: https://quarto.org/docs/presentations/revealjs/ -->

<!-- this, too: https://quarto.org/docs/authoring/markdown-basics.html -->

```{r}
#| include: false
#| echo: false

library(tidyverse)
library(infer)
library(openintro)
```

-   Revisit `cpr` [^1] dataset we explored in Ch. 14
-   2 variables
    -   *group*: treatment (received blood thinner) or control (did not)
    -   *outcome*: died or survived (for at least 24 hours)
-   90 patients (40 treatment, 50 control, randomly assigned)

[^1]: `cpr` is from the *openintro* package

## Hypotheses:

-   $H_0$: Blood thinners do not affect survival rate. $p_T-p_C = 0$
-   $H_A$: Blood thinners affect survival rate. $p_T-p_C \neq 0$

## Data:

| group     | died | survived | total |
|-----------|:----:|:--------:|------:|
| control   |  39  |    11    |    50 |
| treatment |  26  |    14    |    40 |
| total     |  65  |    25    |    90 |

Difference in proportions of "survived": $$\hat{p}_T-\hat{p}_C=\frac{14}{40}-\frac{11}{50}=0.13$$

## Hypothesis Test Using Random Permutation

::: panel-tabset
### Simulations

-   1,000 random permutations simulating true null hypothesis
-   Values of response (outcome) shuffled each time

```{r}
#| include: true
#| echo: true
set.seed(8675390)

library(infer)
cpr_perm <- cpr |>
  specify(outcome ~ group, success = "survived") |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  calculate(stat = "diff in props", order = c("treatment", "control"))
```

### Null Distribution

```{r}
#| include: true
#| echo: false
#| fig-cap: "Null distribution for difference in proportions that survived (treatment - control). Values that are at least as extreme as observed value indicated by dashed lines."

cpr_perm |>
  ggplot(aes(stat)) +
  geom_histogram(bins = 14) +
  geom_vline(xintercept = -0.13,
             color = "red", linewidth = 2, linetype = "dashed") +
  geom_vline(xintercept = 0.13,
             color = "red", linewidth = 2, linetype = "dashed") +
  labs(x = "Difference in randomized survival rates (treatment - control)",
       title = "1,000 differences in randomized proportions") +
  theme_minimal()
```

### p-value

-   p-value = 166/1000 = 0.166

```{r}
#| include: true
#| echo: true

cpr_perm |>
  summarize(count_extreme = sum(abs(stat) >= 0.13),
            pval = mean(abs(stat) >= 0.13))
```
:::

## Mathematical Model for Difference in Proportions

::: callout-note
## Sampling distribution of $\hat{p}_1-\hat{p}_2$

The sampling distribution of $\hat{p}_1-\hat{p}_2$ based on samples of size $n_1$ and $n_2$ and population proportions $p_1$ and $p_2$ will be approximately normal with mean $p_1-p_2$ and standard error $$SE(\hat{p}_1-\hat{p}_2)=\sqrt{\frac{p_1(1-p_1)}{n_1}+\frac{p_2(1-p_2)}{n_2}}$$\
if the following technical conditions are met:

1.  Data are independent within and between the two groups (e.g., observations from two independent random samples or from a randomized experiment)
2.  (**success-failure condition**) At least 10 expected successes and at least 10 expected failures in each group.
:::

## Hypothesis Test Using Normal Approximation

-   Under the null hypothesis $p_1=p_2$
-   We use the **pooled** proportion of successes, $\hat{p}_{pool}$ to approximate this common proportion $$\hat{p}_{pool}=\frac{number\, of\, successes}{number\, of\, cases}=\frac{\hat{p}_1n_1+\hat{p}_2n_2}{n_1+n_2}$$
-   In the CPR example 25 survived out of 90 total cases, so $\hat{p}_{pool}=25/90=0.278$

## Checking Conditions for Hypothesis Test

-   The expected numbers of successes and failures in group 1 are $n_1\hat{p}_{pool}$ and $n_1(1-\hat{p}_{pool})$
-   In group 2 they are $n_2\hat{p}_{pool}$ and $n_2(1-\hat{p}_{pool})$

------------------------------------------------------------------------

In the CPR example we expect

-   Treatment group
    -   $0.278\cdot40=11.1$ successes
    -   $(1-0.278)\cdot40=28.9$ failures
-   Control group
    -   $0.278\cdot50=13.9$ successes
    -   $(1-0.278)\cdot50=36.1$ failures

Since there are at least 10 expected successes and failures in each group a normal approximation of the null distribution is appropriate

## SE for Hypothesis Test Using Normal Approximation

-   We also use the pooled proportion to approximate the SE $$\begin{array}{rcl}SE(\hat{p}_1-\hat{p}_2) & \approx & \sqrt{\frac{\hat{p}_{pool}(1-\hat{p}_{pool})}{n_1}+\frac{\hat{p}_{pool}(1-\hat{p}_{pool})}{n_2}}\\ & = & \sqrt{\hat{p}_{pool}(1-\hat{p}_{pool})\left(\frac{1}{n_1}+\frac{1}{n_2}\right)}\end{array}$$
-   For the CPR study $$SE\approx \sqrt{0.278\cdot(1-0.278)\left(\frac{1}{40}+\frac{1}{50}\right)}=0.095$$

## Z Score for Two Proportions

-   The hypothesis test using a normal approximation uses the $Z$ score as the test statistic $$Z = \frac{(\hat{p}_1-\hat{p}_2) - 0}{\sqrt{\hat{p}_{pool}(1-\hat{p}_{pool})\left(\frac{1}{n_1}+\frac{1}{n_2}\right)}}$$
-   Note that the denominator is the SE estimate we saw in the previous slide
-   When the conditions are met, $Z$ will have a standard normal distribution ($N(0,1)$)

------------------------------------------------------------------------

For the CPR example the Z score is $$Z=\frac{(\hat{p}_T-\hat{p}_C)-0}{SE}=\frac{0.13}{0.095}=1.37$$

## P-value

```{r}
#| include: true
#| echo: false
#| fig-cap: "Standard normal curve with shaded area corresponding to p-value"

normTail(m=0, s=1, L=-1.37, U=1.37)
```

The 2-sided p-value is the area under the density curve for $N(0,1)$ that is more extreme than the observed difference ($\leq-1.37$ or $\geq1.37$)

```{r}
#| include: true
#| echo: true

2*pnorm(-1.37, mean = 0, sd = 1)
```

------------------------------------------------------------------------

Compare this p-value (0.171) to the one we calculated using random permutation (0.166)

## Bootstrap Percentile Confidence Interval

-   We can calculate a bootstrap percentile 95% confidence interval in much the same way that we did for a single proportion in Ch 12
-   We think about the two samples (groups) as being our best approximation of the population and resample with replacement (bootstrap) from each group ($n_1$ from group 1, $n_2$ from group 2)
-   The bootstrap proportions $\hat{p}_{1,boot}$ and $\hat{p}_{2,boot}$ will tend to be centered on $\hat{p}_1$ and $\hat{p}_2$ but will vary between replicates

------------------------------------------------------------------------

-   Calculate difference in bootstrap proportions $\hat{p}_{1,boot}-\hat{p}_{2,boot}$ for each of a large number of replicates (at least 1,000)
-   95% CI is given by 2.5% to 97.5% percentiles

------------------------------------------------------------------------

Let's compute 1,000 differences in bootstrapped proportions using the CPR data.

```{r}
#| include: true
#| echo: true

cpr_boot <- cpr |>
  specify(outcome ~ group, success = "survived") |>
  generate(reps = 1000, type = "bootstrap") |>
  calculate(stat = "diff in props", order = c("treatment", "control"))
```

The 95% bootstrap percentile confidence interval for the difference in survival rates (treatment - control) is between -0.055 and 0.313.

```{r}
#| include: true
#| echo: true

cpr_boot |>
  summarize(ci_lo = quantile(stat, 0.025),
            ci_hi = quantile(stat, 0.975))
```

## Bootstrap SE Confidence Interval

-   Another way to compute a confidence interval is to use the differences in bootstrapped proportions to estimate the standard error
-   For the CPR data this gives us the estimate SE $\approx$ 0.0944

```{r}
#| include: true
#| echo: true

cpr_boot |>
  summarize(se = sd(stat))
```

------------------------------------------------------------------------

-   This gives us a 95% Bootstrap SE Confidence Interval of $$0.13\pm 1.96\cdot 0.0944$$
-   Thus, the 95% confidence interval is between -0.055 and 0.315

## Confidence Interval Using Normal Approximation

-   We can also use a normal approximation to calculate a confidence interval if the technical conditions are met
-   In this case, we use $\hat{p}_1$ and $\hat{p}_2$ as the best approximations of $p_1$ and $p_2$

## Checking Conditions for Using Normal Approximation for CI

-   In this case, the expected numbers of successes and failures in each group are the same as the counts of successes and failures in the samples
-   The CPR data satisfy the success-failure condition

## SE for Using Normal Approximation for CI

```{r}
#| include: false
#| echo: false

pt <- 14/40
pc <- 11/50
nt <- 40
nc <- 50

se <- sqrt(pt*(1-pt)/nt + pc*(1-pc)/nc)
```

-   The standard error approximation is $$SE\approx\sqrt{\frac{\hat{p}_1(1-\hat{p}_1)}{n_1}+\frac{\hat{p}_2(1-\hat{p}_2)}{n_2}}$$
-   For the CPR study $$SE\approx\sqrt{\frac{0.35\cdot(1-0.35)}{40}+\frac{0.22\cdot(1-0.22)}{50}}=0.0955$$

## 95% Confidence Interval

-   Using the normal approximation, the 95% confidence interval for the difference in survival rates is $$0.13\pm 1.96\cdot 0.0955$$
-   Thus, the 95% confidence interval is between -0.057 and 0.317

## Comparison of 95% Confidence Intervals

| Type                 |    Interval     |
|----------------------|:---------------:|
| Bootstrap Percentile | (-0.055, 0.313) |
| Bootstrap SE         | (-0.055, 0.315) |
| Normal Approximation | (-0.057, 0.317) |
