---
title: "Inference: Two-Way Tables"
subtitle: |
  | IMS1 Ch. 18 
  | Math 219
author: "Yurk"
format: 
  revealjs:
    theme: beige
editor: source
---

## Government Spending

<!-- Useful: https://quarto.org/docs/presentations/revealjs/ -->

<!-- this, too: https://quarto.org/docs/authoring/markdown-basics.html -->

```{r}
#| include: false
#| echo: false

library(tidyverse)
library(infer)
library(openintro)
```

```{r}
#| include: false
#| echo: false

gss2016 <- read_rds("data/gss-sampled-2016.rds") |>
  mutate(party = case_when(
    str_detect(partyid, "DEMOCRAT")   ~ "Dem",
    str_detect(partyid, "REPUBLICAN") ~ "Rep",
    str_detect(partyid, "IND")        ~ "Ind",
    TRUE                              ~ "Oth"
  )) |>
  filter(party != "Oth") |>
  droplevels()
```

-   Do people that identify as belonging to different U.S. political parties have different views about government spending?
-   We will explore the relationship between party affiliation and opinions on government spending on both national defense and space exploration

## Hypotheses

::: {style="font-size: 30px"}
**National Defense**

-   $H_0$: There is no difference in opinions on government spending on national defense between people with different political affiliations.
-   $H_A$: There is some difference in opinions on government spending on national defense between people with different political affiliations.

**Space Exploration**

-   $H_0$: There is no difference in opinions on government spending on space exploration between people with different political affiliations.
-   $H_A$: There is some difference in opinions on government spending on space exploration between people with different political affiliations.
:::

## Data:

-   `gss2016` [^1] dataset
-   Subset of General Social Survey (GSS) data from 2016
-   `party` (*Dem*, *Ind*, or *Rep*)
-   `natarms` opinion on current level of government spending on national defense
-   `natspac` opinion on current level of government spending on space exploration
-   149 respondents

[^1]: Available from the [github page for IMS1 Tutorial 5.3](https://github.com/OpenIntroStat/ims-tutorials/tree/master/05-infer/03-lesson/data)

------------------------------------------------------------------------

```{r}
#| include: false
#| echo: false

gss2016 |>
  select(party, natarms) |>
  table()
```

## Military Spending

::: {style="font-size: 30px"}
| Party | TOO LITTLE | ABOUT RIGHT | TOO MUCH | Total |
|-------|:----------:|:-----------:|:--------:|------:|
| Dem   |     17     |     14      |    12    |    43 |
| Ind   |     20     |     28      |    24    |    72 |
| Rep   |     24     |      8      |    2     |    34 |
| Total |     61     |     50      |    38    |   149 |
:::

-   With more than 2 groups, we can't use a single difference in proportions to compare groups
-   We will use the $X^2$ statistic (chi-squared) to measure the difference between groups

------------------------------------------------------------------------

::: {style="font-size: 30px"}
| Party | TOO LITTLE | ABOUT RIGHT |  TOO MUCH  | Total |
|-------|:----------:|:-----------:|:----------:|------:|
| Dem   | 17 (17.60) | 14 (14.43)  | 12 (10.97) |    43 |
| Ind   | 20 (29.48) | 28 (24.16)  | 24 (18.36) |    72 |
| Rep   | 24 (13.92) |  8 (11.41)  |  2 (8.67)  |    34 |
| Total |     61     |     50      |     38     |   149 |
:::

-   First compute the expected cell counts, assuming $H_0$
-   Overall proportion of people that said "too little" spending = 61/149 = 0.4094
-   If no association between party and opinion, we expect 0.4094 proportion of dems to have this opinion
-   Expected count for dems with opinion "too little" = $0.4094\cdot 43 = 17.60$

------------------------------------------------------------------------

::: {style="font-size: 20px"}
| Party |             TOO LITTLE             |            ABOUT RIGHT            |             TOO MUCH              |
|---------------|:-------------------------:|:-------------:|:-------------:|
| Dem   | $\frac{(17 -17.60)^2}{17.60}=0.02$ | $\frac{(14-14.43)^2}{14.43}=0.01$ | $\frac{(12-10.97)^2}{10.97}=0.10$ |
| Ind   | $\frac{(20-29.48)^2}{29.48}=3.05$  | $\frac{(28-24.16)^2}{24.16}=0.61$ | $\frac{(24-18.36)^2}{18.36}=1.73$ |
| Rep   | $\frac{(24-13.92)^2}{13.92}=7.30$  | $\frac{(8-11.41)^2}{11.41}=1.02$  |  $\frac{(2-8.67)^2}{8.67}=5.13$   |
:::

-   Compute $\frac{(observed\,count-expected\,count)^2}{expected\,count}$ for each cell
-   Add values to obtain $X^2$ statistic, $$\begin{array}{rcl}X^2 &=& 0.02+0.01+0.10 \\ & + & 3.05 + 0.61 + 1.73 \\ & + & 7.30 + 1.02 + 5.13 \\ & =& 18.97\end{array}$$

------------------------------------------------------------------------

Or just ask R...

```{r}
#| include: true
#| echo: true

library(infer)
X2_arm <- gss2016 |> 
  observe(natarms ~ party, stat = "Chisq") |> 
  pull()
X2_arm
```

------------------------------------------------------------------------

## Spending on Space Exploration

::: panel-tabset
### Counts

```{r}
#| include: false
#| echo: false

gss2016 |>
  select(party, natspac) |>
  table()
```

| Party | TOO LITTLE | ABOUT RIGHT | TOO MUCH | Total |
|-------|:----------:|:-----------:|:--------:|------:|
| Dem   |     8      |     22      |    13    |    43 |
| Ind   |     13     |     37      |    22    |    72 |
| Rep   |     9      |     17      |    8     |    34 |
| Total |     30     |     76      |    43    |   149 |

### Expected Counts

| Party | TOO LITTLE | ABOUT RIGHT |  TOO MUCH  | Total |
|-------|:----------:|:-----------:|:----------:|------:|
| Dem   |  8 (8.66)  | 22 (21.93)  | 13 (12.41) |    43 |
| Ind   | 13 (14.50) | 37 (36.72)  | 22 (20.78) |    72 |
| Rep   |  9 (6.85)  | 17 (17.34)  |  8 (9.81)  |    34 |
| Total |     30     |     76      |     43     |   149 |

### $X^2$ Statistic

This time we will go straight to R to calculate $X^2$

```{r}
#| include: true
#| echo: true

X2_spac <- gss2016 |> 
  observe(natspac ~ party, stat = "Chisq") |> 
  pull()
X2_spac
```
:::

## Randomization Test for Independence

-   We can randomly permute the response (opinion) to simulate the null hypothesis being true
-   For each permuted sample, we calculate value of the $X^2$ statistic
-   Let's construct a null distribution for the military spending question

```{r}
#| include: true
#| echo: true
set.seed(8675309)

arm_perm <- gss2016 |>
  specify(natarms ~ party) |>
  hypothesise(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  calculate(stat = "Chisq")
```

------------------------------------------------------------------------

```{r}
#| include: true
#| echo: false
#| fig-cap: "Histogram of $X^2$ statistics for 1,000 random permutations. Observed value indicated by dashed vertical line."

arm_perm |>
  ggplot(aes(stat)) +
  geom_histogram() +
  geom_vline(xintercept = X2_arm, color = "red", linewidth = 2, linetype = "dashed") +
  labs(title = "1,000 randomized statistics",
       x = "Chi-squared statistics") +
  theme_minimal()
```

------------------------------------------------------------------------

-   There were no values of $X^2$ that were as extreme as the observed value
-   The p-value is approximately 0
-   We can conclude that opinions on military spending are associated with political party (the two variables are not independent)

```{r}
#| include: true
#| echo: false

arm_perm |>
  summarize(extreme_count = sum(stat >= X2_arm), pval = mean(stat >= X2_arm))
```

## Test for Independence Using a Mathematical Model

::: callout-note
## Chi-squared test for assessing independence between categorical variables

When the null-hypothesis is true and the following conditions are met, $X^2$ has a Chi-squared distribution with $df=(r-1)\times(c-1)$ degrees of freedom:

1.  Independent observations
2.  Large samples: at least 5 expected counts in each cell
:::

-   $r$ is the number of rows and $c$ is the number of columns in the two-way table (no totals)

------------------------------------------------------------------------

-   Both two-way tables satisfy the large samples condition (at least 5 expected counts in each cell)
-   In both cases there are 3 rows and 3 columns in the table, so $df=(3-1)\times(3-1)=4$

```{r}
#| include: true
#| echo: false
#| fig-cap: "Chi-squared disribution with $df=4$. Purple line shows observed value for space question. Red line shows observed value for military question."

x2d1 <- tibble(x = seq(0, 20, by=0.1)) |>
  mutate(y = dchisq(x, df = 4))

x2d1 |>
  ggplot(aes(x, y)) +
  geom_line(linewidth = 1.5) +
  geom_vline(xintercept = X2_arm, color = "red", linewidth = 2, linetype = "dashed") +
  geom_vline(xintercept = X2_spac, color = "purple", linewidth = 2, linetype = "dashed") +
  labs(x = "Chi-squared",
       y = "") +
  theme_minimal()
```

------------------------------------------------------------------------

-   The p-value is the area under the curve that is beyond the observed $X^2$ value
-   The `pschisq` function computes the area up to the specified cutoff, subtract value from 1 to find the p-value
-   Here are the p-values for the two hypothesis tests

**Military Spending**

```{r}
#| include: true
#| echo: true

1 - pchisq(X2_arm, df = 4)
```

**Space Exploration**

```{r}
#| include: true
#| echo: true

1 - pchisq(X2_spac, df = 4)
```

------------------------------------------------------------------------

-   As with the randomization-based test, the p-value is very small (\<0.001) for the military spending question
-   The p-value is quite large ($p=0.857$) for the space exploration space exploration question
-   We cannot reject null hypothesis in the latter case
-   It is plausible that opinion on government spending on space exploration and political affiliation are independent

## $X^2$ distributions for different $df$

```{r}
#| include: true
#| echo: false
#| fig-cap: "Chi-squared disributions with different degrees of freedom."

x2d2 <- tibble(x = seq(0, 20, by=0.1)) |>
  mutate(df_2 = dchisq(x, df = 2),
         df_4 = dchisq(x, df = 4),
         df_9 = dchisq(x, df = 9)) |>
  pivot_longer(cols = starts_with("df"), names_to = "df", names_prefix = "df_", values_to = "y")

x2d2 |>
  ggplot(aes(x, y, color = df, linetype = df)) +
  geom_line(linewidth = 1.5) +
  labs(x = "Chi-squared",
       y = "") +
  theme_minimal()
```

-   Chi-squared distribution is more peaked for lower $df$
-   Thicker tail for higher $df$
