---
title: "Compare Two Independent Means"
subtitle: |
  | IMS1 Ch. 20 
  | Math 219
author: "Yurk"
format: 
  revealjs:
    theme: beige
editor: source
---

## Birth Weights and Smoking

<!-- Useful: https://quarto.org/docs/presentations/revealjs/ -->

<!-- this, too: https://quarto.org/docs/authoring/markdown-basics.html -->

```{r}
#| include: false
#| echo: false

library(tidyverse)
library(infer)
library(openintro)
```

```{r}
#| include: false
#| echo: false
data(births14)

births14 <- births14 |>
  drop_na(habit, weight) |>
  select(habit, weight)
```

-   Do infants whose mothers smoke have a different mean birth weight than infants whose mothers do not smoke?
- Let $\mu_n$ be the mean weight (lbs) of infants whose mothers did not smoke, and let $\mu_s$ be the mean for infants whose mothers smoked

## Inference

-  We will estimate the difference in mean birth weights $\mu_n-\mu_s$ using a confidence interval
-   We will conduct a hypothesis test with hypotheses
    -   $H_0: \mu_n-\mu_s = 0$
    -   $H_A: \mu_n-\mu_s \neq 0$

## Data

-   `births14` [^1] dataset
-   Random sample of 1,000 cases from US birth data set from 2014 (19 removed with missing values)
-   `habit` is smoking habit ("smoker" or nonsmoker")
- `weight` is birth weight in pounds

[^1]: `births14` is from the `openintro` package

## EDA

::: panel-tabset
### Histograms

```{r}
#| include: true
#| echo: false
#| fig-cap: "Histrograms showing birth weights for infants whose mothers did not smoke (top) for infants whose mothers smoked (bottom)."

births14 |>
  ggplot(aes(weight, fill = habit)) +
  geom_histogram(bins = 15, color = "white") +
  labs(x = "newborn weight (lbs)") +
  facet_wrap(vars(habit), ncol = 1, scales = "free_y") +
  theme_minimal() +
  theme(axis.text=element_text(size=16),
        axis.title=element_text(size=14),
        strip.text=element_text(size=16),
        legend.position = "none")
```

### Summary Statistics

```{r}
#| include: false
#| echo: false

b14stat <- births14 |>
  group_by(habit) |>
  summarize(n = n(),
            mean = mean(weight),
            sd = sd(weight))

dif_mean <- b14stat |>
  summarize(dif = -diff(mean)) |>
  pull()
```

| habit |  n  | mean  |  sd   |
|--|:---:|:-----:|:-----:|
| nonsmoker | 867 | 7.27 | 1.23 |
| smoker | 114 | 6.68 | 1.60 |

The observed difference in means is $$\begin{array}{lcr}\bar{x}_n-\bar{x}_s &=& 7.27-6.68\\ &=& 0.59\end{array}$$

:::

## Randomization Test for Difference in Means

- We can simulate a true null hypothesis by randomly permuting the values of the response variable

```{r}
#| include: true
#| echo: true

set.seed(8675309)

library(infer)
births_perm <- births14 |>
  specify(weight ~ habit) |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  calculate(stat = "diff in means", order = c("nonsmoker", "smoker"))
```

---

```{r}
#| include: true
#| echo: false
#| fig-cap: "Histogram of differences in means (null distribution) calculated from 1,000 random permutations of birth weights. Observed difference is indicated by dashed vertical line."

births_perm |>
  ggplot(aes(stat)) +
  geom_histogram(bins = 30, color = "white") +
  geom_vline(xintercept = dif_mean, color = "red", linetype = "dashed", linewidth = 2) +
  labs(x = "difference in randomized means (nonsmoker - smoker)") +
  theme_minimal()
```


---

- The p-value is the proportion of randomized differences that are at least as extreme as the observed value ($\leq$ `r round(-dif_mean,2)` or $\geq$ `r round(dif_mean,2)`)
- There are no such randomized differences, so the p-value is less than 0.001 (1 in 1000)

```{r}
#| include: true
#| echo: true

births_perm |>
  summarize(n_extreme = sum(abs(stat >= dif_mean)),
            pval = mean(abs(stat) >= dif_mean))
```

## Test Statistic for Comparing Two Means

- The test statistic for comparing two means is the $T$ statistic ($T$ score)
- We will use a version of the $T$ statistic that assumes the two populations have equal variance (different than the version presented in the text)

## Pooled Sample Standard Deviation

- First we compute the pooled sample standard deviation,
$$s_p = \sqrt{\frac{(n_1-1)s_1^2+(n_2-1)s_2^2}{n_1+n_2-2}}$$
- The pooled sample standard deviation in birth weights is
$$\begin{array}{rcl} s_p &=&  \sqrt{\frac{(867-1)\cdot 1.23^2+(114-1)\cdot 1.60^2}{867+114-2}}\\ &=& 1.28\end{array}$$
---

- The $T$ statistic is
$$T=\frac{(\bar{x}_1-\bar{x}_2)-0}{s_p\sqrt{\frac{1}{n_1}+\frac{1}{n_2}}}$$
- For the birth weight example, the value is
$$T=\frac{0.59-0}{1.28\cdot\sqrt{\frac{1}{867}+\frac{1}{114}}} = 4.63$$

## Mathematical Model for Testing the Difference in Means

::: callout-note

When the null hypothesis is true and the following conditions are met, the $T$ score has a $t$-distribution with $df=n_1+n_2-2$ degrees of freedom.

1. Groups have equal variance in the population 
2. Independent observations within and between groups
3.  Normality: Large samples and no extreme outliers.
:::

## Two Sample T-Test

- The degrees of freedom for the birth weight example is $df=867+114-2=979$.
- The p-value is extremely small

```{r}
#| include: true
#| echo: true

2*pt(-4.63, df = 979)
```

## Note About Relaxing the Equal Variance Assumption

- We can compute a $T$ statistic without assuming equal variances (see the formula in the book)
- If the null hypothesis is true and the technical conditions are met, then the distribution of these $T$ statistics will be be *approximately* $t$-distributed 
- The $df$ for the approximating $t$ distribution involves a complicated calculation (not the one listed in the text)

---

- We can use the `t_test` function in the `infer` package to calculate a p-value
- It calculates the $T$ statistic, $df$, and the p-value for us
- It does **NOT** check conditions
- If we specify the option `var.equal = TRUE`, these calculations will use the equal variance assumption

---

- The value of $T$ and the p-value differ from ours due to rounding

```{r}
#| include: true
#| echo: true
births14 |> 
  t_test(weight ~ habit,
         var.equal = TRUE,
         order = c("nonsmoker", "smoker"))
```

---

- Specifying `var.equal = TRUE` relaxes the equal variance assumption
- Note that $df$ is no longer an integer

```{r}
#| include: true
#| echo: true
births14 |> 
  t_test(weight ~ habit,
         var.equal = FALSE,
         order = c("nonsmoker", "smoker"))
```


## Bootstrap Confidence Interval for Difference in Means

- The method for calculating bootsrap confidence intervals for a difference in means is similar to the methods we used for a difference in proportions
- We create bootstrap samples from each group (resampling with replacement) and calculate a difference in means
- We do this 1,000 time and use the resulting sampling distribution to calculate a bootstrap percentile CI or a boostrap SE CI

---

- Lets calculate differences in bootstrapped means for the birth weight example
- The 95% boostrap percentile CI is (0.312, 0.917)

```{r}
#| include: true
#| echo: true

births_boot <- births14 |>
  specify(weight ~ habit) |>
  generate(reps = 1000, type = "bootstrap") |>
  calculate(stat = "diff in means", order = c("nonsmoker", "smoker"))

births_boot |>
  summarize(ci_lo = quantile(stat, 0.025),
            ci_hi = quantile(stat, 0.975))
```

## Estimating the Difference in Means Using a Mathematical Model

- If the technical conditions are met, including the equal variance assumption, then we can use the $t$-distribution to estimate the difference in means
- We can calculate a confidence interval for the difference in means as $$(\bar{x}_1-\bar{x}_2)\pm t^{\ast}_{df}\times SE$$

---

- Assuming equal variance, $df=n_1+n_2-2$, and the standard error is $$SE = s_p\sqrt{\frac{1}{n_1}+\frac{1}{n_2}}$$
- The value of $t^{\ast}_{df}$ depends on the degrees of freedom and the confidence level
- For the birth weights example $$SE=1.28\cdot\sqrt{\frac{1}{867}+\frac{1}{114}}=0.128$$

---

- Since $df=979$ for this example, the value of $t^{\ast}_{df}$ for a 95% confidence interval is 1.96
- Thus, the 95% confidence interval is $$0.59\pm1.96\times0.128=0.59\pm0.251$$
- In interval form it is approximately (0.339, 0.841)

```{r}
#| include: true
#| echo: true

qt(0.975, df = 979)
```

## Relaxing the Equal Variance Assumption for CI

- We can also use the `t_test` function to calculate CI
- Here is the version with the equal variance assumption

```{r}
#| include: true
#| echo: true

births14 |> 
  t_test(weight ~ habit,
         var.equal = TRUE,
         conf_int = TRUE,
         conf_level = 0.95,
         order = c("nonsmoker", "smoker"))
```

---

- Here is the confidence interval calculated without assuming equal variances

```{r}
#| include: true
#| echo: true

births14 |> 
  t_test(weight ~ habit,
         var.equal = FALSE,
         conf_int = TRUE,
         conf_level = 0.95,
         order = c("nonsmoker", "smoker"))
```

## Conclusions

- We get the same result from both versions of the hypothesis test.
- We reject the null hypothesis at the $\alpha=0.05$ significance level and conclude that there is strong evidence of a difference in the average weights of infants born to mothers who smoked and those who did not (p < 0.001)

---

- We are 95% confident that the mean weight of babies born to mothers who did not smoke is between 0.285 and 0.900 pounds higher than the mean weight of babies whos mothers smoked
- This result is also consistent with the result of the hypothesis test, since 0 does not appear in the 95% confidence interval