---
title: "Observational Studies and Stratification"
subtitle: |
  | Additional Topic 
  | Math 215
author: "Yurk"
format: 
  revealjs:
    theme: beige
editor: source
---

```{r}
#| include: false
#| echo: false

library(tidyverse)
library(infer)
library(openintro)
library(broom)
library(HH)
```

## Reducing Unexplained Variability

- The primary role of blocking in experiments is to reduce the unexplained variability in the response
- **Stratifying** plays a similar role in observational studies
- Groups of observational units are formed with similar values of the **stratifying variable**
- Stratifying variable is accounted for as a source of variability in the analysis (similar to blocking in experiment)


## Confounding Variables

- **Confounding variables** are associated with both the dependent variable and the independent variable
- Confounders can obscure the true relationship between the variables of interest
- In experiments, randomization ensures that treatment groups are similar in terms of confounding variables
- This is not the case in observational studies
- Often, the stratifying variable is a potential confounding variable

<!-- ## kjklj -->


<!-- ```{r} -->
<!-- lm(income ~ race, data = acs12) |> -->
<!--   anova() |> -->
<!--   tidy() -->
<!-- ``` -->



<!-- ```{r} -->
<!-- lm(income ~ edu + race, data = acs12) |> -->
<!--   anova() |> -->
<!--   tidy() -->
<!-- ``` -->

<!-- ```{r} -->
<!-- lm(income ~ race + edu, data = acs12) |> -->
<!--   anova() |> -->
<!--   tidy() -->
<!-- ``` -->


<!-- ## Smoking and birth weight -->

<!-- ```{r} -->
<!-- lm(weight ~ smoke, data = births) |> -->
<!--   anova() |> -->
<!--   tidy() -->
<!-- ``` -->

<!-- ```{r} -->
<!-- lm(weight ~ premature + smoke, data = births) |> -->
<!--   anova() |> -->
<!--   tidy() -->
<!-- ``` -->

<!-- ```{r} -->
<!-- lm(weight ~ smoke + premature, data = births) |> -->
<!--   anova() |> -->
<!--   tidy() -->
<!-- ``` -->

<!-- ```{r} -->
<!-- lm(weight ~ weeks + smoke, data = births) |> -->
<!--   anova() |> -->
<!--   tidy() -->
<!-- ``` -->

## Smoking and birth weight

```{r}
#| include: FALSE
#| echo: FALSE

data(births14)

births14 <- births14 |>
  drop_na(habit, weight, weeks, premie)
```

-   Do infants whose mothers smoke have a different mean birth weight than infants whose mothers do not smoke?
-   `births14` [^1] dataset
-   Random sample of 1,000 cases from US birth data set from 2014 (19 removed with missing values)
-   `habit` is smoking habit ("smoker" or "nonsmoker")
- `weight` is birth weight in pounds
- `premie` birth is premature (premie) or full-term

[^1]: `births14` is from the `openintro` package

---

- First, we will include smoking habit as a single explanatory variable (no stratifying variable)
- Although it would be appropriate to compare means for smoking and non-smoking mothers using a t-test (see Ch 20 slides), an F-test gives us the same results

```{r}
#| include: TRUE
#| echo: TRUE

lm(weight ~ habit, data = births14) |>
  anova() |>
  tidy()
```

## `premie` as a Stratifying Variable

- The length of the pregnancy is expected to explain some of the variation in birth weight
- It also has the potential to be a confounding variable if there is an association between smoking and length of pregnancy
- We will use `premie` as a stratifying variable in our analysis

## Statistical Model with a Stratifying Variable

Statistical model for the $k$th observation in group $i$ and stratum $j$,

$$y_{ijk}=\mu+\alpha_i+\beta_j+\varepsilon_{ijk}$$

- $\mu$ is the overall mean
- $\alpha_i$ is the differential effect of group $i$
- $\beta_j$ is the differential effect of stratum $j$
- $\varepsilon_{ijk}\sim N(0,\sigma^2)$ is the noise

## ANOVA Table

- $SS_{premie}$ accounts for the effect of `premie`
- $SS_{habit}$ accounts for both variables, but then $SS_{premie}$ is subtracted off

```{r}
#| include: TRUE
#| echo: TRUE

lm(weight ~ premie + habit, data = births14) |>
  anova() |>
  tidy()
```

---

- Interestingly, $p$-value for `habit` is higher when accounting for `premie` in the model
- This is the opposite of what we would expect to see in an experiment, and is due to association between `premie` and `habit`

```{r}
#| include: TRUE
#| echo: TRUE

lm(weight ~ premie + habit, data = births14) |>
  anova() |>
  tidy()
```

---

We can also see the effects of this association if we compare the coefficient of the corresponding linear models.

```{r}
#| include: TRUE
#| echo: TRUE

lm(weight ~ habit, data = births14) |>
  tidy()
```

`habit` has a smaller effect when we account for `premie`

```{r}
#| include: TRUE
#| echo: TRUE

lm(weight ~ premie + habit, data = births14) |>
  tidy()
```


## Association between `habit` and `premie`

```{r}
#| include: TRUE
#| echo: FALSE

births14 |>
  ggplot(aes(habit)) +
  geom_bar(aes(fill = premie), position = position_fill()) +
  labs(y = NULL) +
  theme_minimal()
```

---

- Due to the association between `habit` and `premie` it is difficult to fully separate their effects on birth weight
- E.g., whether the mother smokes or not *could* affect whether the baby is a premie or not which could impact weight
- $SS_{habit}$ measures the variation in the response that is attributed to `habit` but does not include the variation that is jointly attributed to the two explanatory variables

---

- If we reverse the order of the variables, we can also measure the variation that is attributed to `premie` without variation that is jointly attributed to the two variables

```{r}
#| include: TRUE
#| echo: TRUE

lm(weight ~ habit + premie, data = births14) |>
  anova() |>
  tidy()
```



## ANCOVA

- In the previous analysis we used `premie` as a stratifying variable
- The dataset also includes the variabe `weeks`, the length of the pregnancy in weeks
- We can conduct an alternative analysis using `weeks` as a (numeric) covariate instead of stratifying by `premie`
- Statistical model (parallel lines) for the $j$th observation in the $i$th group:
$$y_{ij}=\mu + \alpha_i + \beta(X_{ij} - \bar{\bar{X}})+\varepsilon_{ij}$$
---

- `weeks` as a covariate explains more of the variation in the response than `premie` ($SS_{weeks} = 483$ vs. $SS_{premie}=374$)
- The resulting model explains more of the variability in the response than the model that included `premie`, so $SSE$ is smaller
- As a result, the $F$ statistic is larger, and the $p$-value is smaller

```{r}
#| include: TRUE
#| echo: TRUE


lm(weight ~ weeks + habit, data = births14) |>
  anova() |>
  tidy()
```

## Conclusions

- Both analyses lead us to conclude that there is a significant association between smoking and birth weight when we also account for the length of pregnancies
- We reach the same conclusion whether we treat the length of pregnancy as a categorical variable (`premie`) or a numeric variable (`weeks`)
- Because this is an observational study, we cannot conclude that smoking causes a difference in birth weight
- It is possible that there are other confounding variables that we have not accounted for in the model